### 2021-05-14 restricted_randomisation.tex
First draft (s.t. examples)

 %%% Multi-stratum designs, MSE-criteria
In this chapter we will adapt the MSE-based criteria derived before to the factorial experiments with units distributed in several strata. First we shall briefly review the approaches to the inference from the experiments with multi-level unit structures, and then will define the MSE-based optimality for the multistratum framework and provide the implementation strategy for the corresponding design search.

In Section \ref{sec::ch7_ex} a few illustrative examples will be considered, for the main purposes of assessing the obtained optimal designs' performances and structures, and investigating the presence of any patterns and tendencies regarding the relationships between the criteria form, parameters and the results.

\section{Response-Surface Methodology for Multistratum Designs}
%% Model. Ratio of variances.
As randomisation occurs at each stratum, the model accounting for the hierarchical error structure and correlated observations can be written as
\begin{equation}
\label{eq::MS_model}
\bm{Y}=\bm{X}\bm{\beta}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{Y}$ represents the $n$-dimensional vector of responses, $\bm{X}$ stands for the $n \times p$ model matrix and $\bm{\beta}$ for the $p$-dimensional vector of corresponding model coefficients. Each row of the $n\times m_{i}$ matrix $\bm{Z}_{i}$ corresponds to a single experimental run and indicates the unit in stratum $i$ containing this run. $\bm{\varepsilon}_{i}$ is a vector of random effects occurring due to the randomisation at level $i$, and these effects are assumed to be independent and identically distributed around zero mean and variance $\sigma^{2}_{i}$. The assumption of error normality, though quite common is required for some analysis methods, is not a necessary one at this stage.

The majority of purposes of inference rely on obtaining parameter estimates $\bm{\hat{\beta}}$ of good quality, that is minimising the uncertainty of the obtained estimates, which implies the necessity of evaluating the dispersion parameters $\sigma^2_{i}$ that are directly related to the estimates' variance and, therefore, responsible for measuring the uncertainty of the inference in a general sense.

\section{REML Methodology}
An extensive amount of research has been conducted on the design of and analysis of data from split-plot and split-split-plot experiments, with one of the first works by \cite{Letsinger1996BiRandomization}, emphasising the necessity of adapting the design and analysis strategy with respect to the more complicated error structure.  In the most general case of non-orthogonal unit structures Residual Maximum Likelihood (REML) methodology has been acknowledged to be the most sensible approach to estimating the variance components by maximising the part of the likelihood function corresponding to the ``random'' part of the model; the details are provided, for example, in \cite{Searle2001generalized}.

In traditional REML-based methodology the estimates of $\sigma^{2}_{i}$ are obtained from model (\ref{eq::MS_model}) and then GLS estimators of $\bm{\beta}$ are calculated as
\begin{equation}
\label{eq::MS_GLS}
\bm{\hat{\beta}}=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}\bm{X}'\bm{V}^{-1}\bm{Y},
\end{equation}
where $\bm{V}$ is the unknown variance-covariance matrix of the response vector, to which the estimates are to be plugged in:
\begin{equation*}
\bm{V}=\sum_{i=1}^{s}\sigma^{2}_{i}\bm{Z}_{i}\bm{Z}'_{i},
\end{equation*}
so that $\mbox{Var}(\bm{\hat{\beta}})=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}.$

However, in some cases REML tends to underestimate the variance components in the higher stratum, for example \cite{Goos2006practical} considered some split-plot experiments with true whole-plot variances being non-negligibly far from zero, for which REML, however, provided zero estimates.

A possible alternative would be a Bayesian analysis, that is specifying prior information regarding the variance components (as was suggested by \cite{Gilmour2009analysis}), however, this requires a careful choice of the prior which is often not possible, especially in the case of more than two strata.

\section{Full Treatment Model}

In the context of model uncertainty it would be appropriate to loosen the assumption of the response surface model (\ref{eq::MS_model}) and consider instead the relationship of the response with the set of treatments:
\begin{equation}
\label{eq::treatmentMS}
\bm{Y}=\bm{X_t}\bm{\tau}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i}.
\end{equation}

Each column of $\bm{X_t}$ here corresponds to a treatment defined as a combination of the experimental factors;
%in case of factorial experiment the total number of possible treatments (or candidate design points) $T$ would be the product of the number of the factors' levels. 
$\bm{Z}_{i}$, as before, is the indicator matrix of random effects at stratum $i$.

\subsection{Pure-error REML}
\cite{GilmourGoos2016PEREML} introduced the notion of ``Pure Error REML'', such that the random effects are estimated from the treatment model (\ref{eq::treatmentMS}), and then for the sake of making inference regarding parameters of the response model (\ref{eq::MS_model}), they are substituted in the GLS estimators of the fixed effects (\ref{eq::MS_GLS}). It is also argued that some appropriate corrections are to be adapted and applied to the obtained estimates, i.e.~the ones suggested by \cite{kenward1997small}.

@@

Pure error is expected to measure the variability between experimental units regardless of the treatments applied, and therefore, the assumption of treatment-unit additivity is essential in unblocked experiments.  In the presence of one or several blocking factors, the existence of block-treatment interaction would imply that pure error in the lower strata can only be estimated from the replicates within blocks, so that inter-block information is not taken into consideration. Hence, in the blocked cases the assumption of treatment-block additivity \citep{Draper1998} is desirable. \cite{Gilmour2000PErsm} argue for the preservation of treatment-unit additivity when either fixed or random block effects are in the model in order to preserve consistency with the unblocked experiments; that would also conveniently imply treatment-block additivity. Adoption of these assumptions would then lead to the following representation of the relationship between responses and treatments in multi-stratum experiments.

@@

Based on the interrelation of the two models, and the desirability of being able to provide the possibility for testing for the lack-of-fit together with obtaining robust estimates of the variance components, the approach of using pure error is adopted, \cite{GilmourGoos2016Robust} explored several approaches to estimating the variance components.

\subsection{Yates' procedure}
\label{sec::ch7_varest}
The first approach is stratum-by-stratum data analysis, where each randomisation level is considered separately, and at each stratum $i$ units are treated as runs aggregated in $m_{i-1}$ blocks with fixed effects. The lower stratum variance is estimated from within blocks (using intra-block residual mean square), and the higher stratum (inter-block) variance from the difference between the intra- and inter-block residual mean squares scaled by the number of runs per block \citep{Hinkelmann2005Advanced}. This simple strategy requires no other assumptions except for the treatment-unit additivity and provides unbiased treatment effects' estimators.

However, as mentioned before, it would be desirable to make use of combining inter- and intra-block information, especially in the common cases of relatively small experiments when the amount of runs available does not allow for obtaining higher stratum variance components from only considering blocks as whole units. Yates' procedure, first suggested by \cite{yates1939recovery} and described in detail by \cite{Hinkelmann2005Advanced}, provides more degrees of freedom for the variance estimate in the higher stratum.

Within-block variance is obtained from the residual mean square, as before. The between-block variation is estimated form fitting the block-after-treatment model: the treatment sum of squares  accounted for the blocks corresponding to the higher strata and the residual sum of squares obtained in the intra-block analysis are subtracted from the total sum of squares. What remains is the extra sum of squares for the blocks in the current stratum, which is used for estimating the intra-block variation.
%Whereas between-block variation is estimated from what is left for the blocking components sum of squares in the current stratum after the sums of squares corresponding to the block effects of the higher strata and all treatments are subtracted from the total sum of squares accounted for the residual sum of squares. 
Schematically these two steps can be presented as below (with the order of the components being crucial):
\begin{enumerate}
\item Total SS = SS(Blocks) + SS(Treatments$\vert$Blocks) + SS(Residual)\\
From fitting this full treatment-after-blocks model the residual mean square $S^2$ is taken as the estimate of the intra-block variance. SS(Residual) is then substituted in the following partition.
\item Total SS = SS(Treatments) + SS(Blocks$\vert$Treatments) + SS(Residual)\\
Therefore, the sums of squares for the blocks after the treatment effects have been accounted for is obtained and the corresponding mean square $$S^2_{b}=\mbox{SS(Blocks$\vert$Treatments)}/\nu_{b}$$ is then used to get an estimate of the inter-block variance:
\item $\hat{\sigma}^2_{b}=\frac{\nu_{b}(S^2_{b}-S^2)}{d}$,\\
where $d=n-\mbox{trace}[\bm{Z}'\bm{X_t}(\bm{X_t}'\bm{X_t})^{-1}\bm{X_t}'\bm{Z}]$, $\nu_{b}=\mbox{rank}([\bm{X_t} \bm{Z}])-\mbox{rank}(\bm{X_t})$, and $\bm{X_t}$ and $\bm{Z}$ are as in (\ref{eq::treatmentMS}).
\end{enumerate}

\cite{GilmourGoos2016Robust} show that in addition to the treatment-unit additivity and randomisation the assumption of the experimental units being a random sample from an infinite population is necessary for the inter-block variance estimate to be unbiased. In this particular context, when the normality and independence of responses in (\ref{eq::MS_model}) is a standard assumption, although a potential presence of contamination effects in the fixed part of the model is also accounted for, Yates' approach seems to be the most appropriate technique. However, relying on the distribution assumption to the extent that makes REML the most sensible approach means that the fixed part of the fitted model is assumed to be absolutely true, that is the parameters of the population distribution are known, which does not comply with the model uncertainty framework.

@@

\section{Design Construction for MSE-based Optimality Criteria}
\label{sec::ch7_search}
With the theoretical insights discussed above in mind, the practical adaptation and implementation of the MSE-based criteria for multistratum experiments shall be presented as the algorithm of the corresponding optimal design search which is in accordance with the inference strategy described in the previous section.

Recall the general outline of a multistratum experiment (as in Section \ref{sec::back_ms}): 
\begin{itemize}
\item There are $s$ strata nested within each other, stratum number $0$ denoting the whole experiment. A higher stratum number corresponds to the lower level in the hierarchical unit structure.
\item Every unit at stratum $i-1$ contains $n_i$ units at the lower stratum $i$, so that stratum $j$ contains $m_j=\prod_{i=1}^{j}n_{i}$ units and in total there are $n=m_{s}=\prod_{i=1}^{s}n_{i}$ runs. 
\item At least one factor is applied at some of the strata, if there is a stratum with no factor applied, then at this level only blocking is accounted for.
\end{itemize}

In the presence of potential model disturbance which is expressed as a matrix of polynomial terms, the full model for a multistratum experiment is then:
\begin{equation}
\label{eq::MS_model_full}
\bm{Y}=\bm{X}_1\bm{\beta}_1+\bm{X}_2\bm{\beta}_2+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{X}_1$ is the primary terms matrix, which was denoted as $\bm{X}$ in the fitted model (\ref{eq::MS_model}).

\subsection{Construction procedure}
The design construction will follow the approach developed by \cite{Trinca2015improved}: an iterative, stratum-by-stratum algorithm, that treats the higher stratum units as fixed blocks; such an approach does not require any prior assumptions regarding the values of the variance components.

@@

Before moving to a more detailed presentation of the design construction algorithm, recall the MSE-based criteria (\ref{eq::MSE_D}), (\ref{eq::MSE_L}), (\ref{eq::MSE_D_B}) and (\ref{eq::MSE_L_B}) that are to be used:

For an unblocked experiment:
\begin{itemize}
\item Determinant-based $MSE(D)$ criterion:
\begin{align}
\label{eq::MSE_D_ms}
\mbox{minimise }&\left[\left|\bm{X}'_{1}\bm{X}_{1}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{X}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
\end{align}
\item Trace-based $MSE(L)$ criterion:
\begin{align}
\label{eq::MSE_L_ms}
\mbox{minimise} &\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{X}_{1})^{-1}F_{1,d;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag\\& \left[\frac{1}{q}\mbox{trace}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times 
\notag\\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{X}_{1})^{-1}+\tau^2\bm{A}\bm{A}'\}\right]^{\kappa_{MSE}}_{.}
\end{align}
\end{itemize}

For a blocked experiment:
\begin{itemize}
\item Determinant-based $MSE(D)$ criterion:
\begin{align}
\label{eq::MSE_D_B_ms}
\mbox{minimise }&\left[\left|(\bm{X}'_{1}\bm{Q}\bm{X}_{1})^{-1}\right|^{1/p}F_{p,d_B;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{QX}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{QX}_1\bm{\tilde{M}}^{-1}\bm{X}_1^{'}\bm{QX}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
\end{align}
\item Trace-based $MSE(L)$ criterion:
\begin{align}
\label{eq::MSE_L_B_ms}
\mbox{minimise }&\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{Q}\bm{X}_{1})^{-1}F_{1,d_B;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag \\&\left[\frac{1}{q}\mbox{trace}\left(\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag \\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{QX}_{1})^{-1}+\tau^2[\bm{\tilde{A}}\bm{\tilde{A}}']_{22}\}\right]^{\kappa_{MSE}}_{.}
\end{align}
\end{itemize}

@@

\subsection{Allocation of degrees of freedom}

The distribution of the degrees of freedom among different components at each stratum of the resulting design needs to be evaluated in accordance with the variance estimation procedure (as described in Section \ref{sec::ch7_varest}), so that both intra- and inter-block information is taken into account.

We shall consider here the case of a split-split-plot experiment (i.e. with $3$ strata), but this strategy is straightforwardly extended to the cases with any number of randomisation levels. At the first $2$ strata, where experimental units are to be further expanded, and treatments applied at these levels are to be replicated for all sub-units and runs within then the available degrees of freedom (after fitting the model at this current level) are split between three components: pure error, inter-plot (inter-whole-plot and inter-sub-plot for the $1$st and $2$nd strata respectively) and lack-of-fit. In the lowest stratum there are, as in the unblocked cases and experiments with fixed block effects, only pure error and lack-of-fit components.

Starting with presenting the way of calculating the number of pure error degrees of freedom at each level, we then proceed to the details of the inter-plot and lack-of-fit degrees of freedom evaluation accompanied by the corresponding R code. In the end of this section we will consider an illustrative example of a design and see how degrees of freedom are allocated.

If we consider an experiment with $n_1$ units in the first stratum, with each of them having $n_2$ units of stratum $2$, and with $n_3$ runs nested within each of those, there are $n=n_1\times n_2\times n_3$ runs in total. The notation to be used is as follows:
\begin{itemize}
\item $\bm{y}$ -- an $n$-dimensional dummy response vector (any randomly generated vector), which is used for fitting a linear model;
\item $\bm{B_1}$ -- an $n$-dimensional vector, such that its $i$-th element is the label of the first stratum unit that contains the $i$-th run. Similarly, $\bm{B_2}$ contains labels of the second stratum units;
\item  $\bm{T_{W}}$ -- a vector of labels for treatments applied at the first stratum (whole-plots); $\bm{T_{S}}$ -- a vector of labels for treatments applied at the second stratum units (sub-plots); this includes treatments in the higher stratum. And, finally, $\bm{T_{SS}}$ contains labels for all treatments applied at the lowest stratum, i.e. at individual runs;
\item $\bm{Xw.m}$, $\bm{Xs.m}$ and $\bm{Xss.m}$ are the matrices containing fixed model terms that are to be fitted at the $1$st, $2$nd and $3$rd strata respectively.
\end{itemize}

If we denote by $p_i$ the number of the fixed model terms fitted at the $i$-th stratum, then the number of available degrees of freedom at each level is: $n_1-1-p_1$ for the $1$st, $n_1\times(n_2-1)-p_2$ -- for the $2$nd, and $n_1\times n_2\times(n_3-1)-p_3$ -- for the $3$rd stratum.

After a design has been obtained, and all the treatment labels, block indicators and model matrices are in place, fitting the following linear models and studying their summary (i.e.~ANOVA) will provide the numbers of degrees of freedom corresponding to every component.

We start with fitting the full treatment model schematically presented in (\ref{eq::model_pe}): the effects of treatments applied to all strata, followed by the terms containing strata indicators, `$\bm{B_i}$'. In the resulting summary we then will see the distribution of pure error degrees of freedom: in the row corresponding to `$\bm{B_1}$' for the whole-plots ($1$st stratum), and in the row corresponding to `$\bm{B_2}$' for the sub-plots ($2$nd stratum). The `Residual' number of degrees of freedom is equal to the number of pure error degrees of freedom for the $3$rd stratum.
\begin{equation}
\label{eq::model_pe}
\mathrm{lm}(\bm{y}\sim\bm{T_{SS}}+\bm{B_1}+\bm{B_2}).
\end{equation}
The order of terms in the expression above is crucial: this is a block-after-treatment model fitting; and the pure error degrees of freedom for lower-level plots occur not only when they are replicated within the same higher-level plot, but also when some pairs of treatments are split between two (or more) higher-level plots. An example of a design is given at the end of this section together with a detailed description of how each pure error degree of freedom is obtained.

From the summary of model (\ref{eq::model_pe}) we automatically obtain the number of lack-of-fit degrees of freedom (LoF) for the lowest stratum, as $n_1\times n_2\times(n_3-1)-p_3 = $`Residual' P.E. $+$ LoF.

>>>>

\textit{	
	\begin{itemize}
		\item[-] General framework of multistratum experiments. [add more on model uncertainty]
		\item[-] Error estimation: REML and pure error REML. Yates procedure
		\item[-] Adapting optimality criteria and design construction
	\end{itemize}
}

In a large number of industrial, engineering and laboratory-based experiments, either the nature of the study or certain restrictions result in the necessity of considering a multi-level structure of experimental units. For example, a chemical process consisting of applying treatments to the material batches of different sizes at each stage;  or one or several experimental factors' values can be changed only once per a certain amount of runs whereas values of other factors are varied between runs. Therefore, different factors are applied at different levels, and randomisation is performed at each level, thus the whole process of allocating treatments to experimental units should be amended accordingly \citep{MeadGilmour2012}.

In this work we go on to consider experimental framework comprising such a hierarchical structure of experimental units and treatments, which is referred to as multistratum experiment, and each stratum is defined as a level in the unit structure. Units are grouped into whole-plots, each of them divided into sub-plots, which contain a certain number of sub-sub-plots, and so on up to the smallest units --- runs of the experiment -- [Figure .with Hasse diagram?] . In the general case of two strata, we deal with what is called a ``split-plot'' experiment, in case of three strata a ``split-split-plot'' experiment. 
%\begin{figure}[h]
%	\begin{center}
%		\includegraphics[scale=0.85]{Hasse.jpg}      %width=\textwidth
%		\caption{Hasse diagram for factors}
%		\label{Fig::Hasse}
%	\end{center}
%\end{figure} 
Our aim is to adapt the MSE-based criteria derived before to the factorial experiments with units distributed in several strata. The most important aspects are (1) estimating the error components at all of the levels and (2) formulating the compliant criteria and provide the suitable implementation strategy for the optimal design search.

\subsection{Response-Surface Methodology for multistratum framework}
%% Model. Ratio of variances.
[OE: introduce the notion of "block"?]
Here we are considering experiments with $s$ strata in total, stratum $i$ being nested within units of the stratum $i-1$, and stratum $0$ will be seen as the whole experiment (following the notation set by \citet{Trinca2015improved}). The number of units in stratum $i$ within every unit in stratum $i-1$ is denoted by $n_i$, such that $m_{j}=\prod_{i=1}^{j}n_{i}$ is the number of units in stratum $j$ and, therefore,  $n=m_{s}=\prod_{i=1}^{s}n_{i}$ is the total number of runs. As randomisation occurs at each stratum, the model accounting for the hierarchical error structure and correlated observations can be written as
\begin{equation}
\label{eq::MS_model}
\bm{Y}=\bm{X}\bm{\beta}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{Y}$ represents the $n$-dimensional vector of responses, $\bm{X}$ stands for the $n \times p$ model matrix and $\bm{\beta}$ for the $p$-dimensional vector of corresponding model coefficients. Each row of the $n\times m_{i}$ matrix $\bm{Z}_{i}$ corresponds to a single experimental run and indicates the unit in stratum $i$ containing this run. $\bm{\varepsilon}_{i}$ is a vector of random effects occurring due to the randomisation at level $i$, and these effects are assumed to be independent and identically distributed around zero mean and variance $\sigma^{2}_{i}$. Generalised Least Squared  (GLS) estimators of $\bm{\beta}$ are calculated as
\begin{equation}
\label{eq::MS_GLS}
\bm{\hat{\beta}}=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}\bm{X}'\bm{V}^{-1}\bm{Y}.
\end{equation}

%The assumption of error normality, though quite common is required for some analysis methods, is not a necessary one at this stage.

%Inferential purposes of obtaining parameter estimates $\bm{\hat{\beta}}$ of good quality imply the necessity of evaluating the dispersion parameters $\sigma^2_{i}$ that are directly related to the estimates' variance and, therefore, responsible for measuring the uncertainty of the inference in a general sense.

\subsubsection*{Error Estimate}
An extensive amount of research has been conducted on the design of and analysis of data from split-plot and split-split-plot experiments, with one of the first works by \cite{Letsinger1996BiRandomization}, emphasising the necessity of adapting the design and analysis strategy with respect to the more complicated error structure.  In the most general case of non-orthogonal unit structures Residual Maximum Likelihood (REML) methodology has been acknowledged to be the most sensible approach to estimating the variance components by maximising the part of the likelihood function corresponding to the ``random'' part of the model; the details are provided, for example, in \cite{Searle2001generalized}.

%In traditional REML-based methodology the estimates of $\sigma^{2}_{i}$ are obtained from model (\ref{eq::MS_model}) and then GLS estimators of $\bm{\beta}$ are calculated as
%\begin{equation}
%\label{eq::MS_GLS}
%\bm{\hat{\beta}}=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}\bm{X}'\bm{V}^{-1}\bm{Y},
%\end{equation}
%where $\bm{V}$ is the unknown variance-covariance matrix of the response vector, to which the estimates are to be plugged in:
%\begin{equation*}
%\bm{V}=\sum_{i=1}^{s}\sigma^{2}_{i}\bm{Z}_{i}\bm{Z}'_{i},
%\end{equation*}
%so that $\mbox{Var}(\bm{\hat{\beta}})=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}.$

In some case, though,  REML tends to underestimate the variance components in the higher stratum, for example \cite{Goos2006practical} considered some split-plot experiments with true whole-plot variances being non-negligibly far from zero, for which REML, however, provided zero estimates. A possible alternative of following Bayesian strategy \citep{Gilmour2009analysis}, however, this requires a careful choice of the prior which is often not possible, especially in the case of more than two strata.

In the context of model uncertainty, it would be appropriate to consider instead the direct relationship between the response and the set of treatments, similar to (\ref{eq::treatment_model}):
\begin{equation}
\label{eq::treatmentMS}
\bm{Y}=\bm{X_t}\bm{\mu_t}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i}.
\end{equation}

Each column of $\bm{X_t}$ corresponds to a treatment defined as a combination of the experimental factors;
%in case of factorial experiment the total number of possible treatments (or candidate design points) $T$ would be the product of the number of the factors' levels. 
$\bm{Z}_{i}$ is the indicator matrix of random effects at stratum $i$.

\cite{GilmourGoos2016PEREML} introduced the notion of ``Pure Error REML'', where random effects are estimated from the treatment model (\ref{eq::treatmentMS}), and then for the sake of making inference regarding parameters of the response model (\ref{eq::MS_model}), they are substituted in the GLS estimators of the fixed effects (\ref{eq::MS_GLS}). It is also argued that some appropriate corrections are to be adapted and applied to the obtained estimates, i.e.~the ones suggested by \cite{kenward1997small}.

@@

[OE: The following is to be paraphrased -- made much shorter or removed overall?]
Pure error is expected to measure the variability between experimental units regardless of the treatments applied, and therefore, the assumption of treatment-unit additivity is essential in unblocked experiments.  In the presence of one or several blocking factors, the existence of block-treatment interaction would imply that pure error in the lower strata can only be estimated from the replicates within blocks, so that inter-block information is not taken into consideration. Hence, in the blocked cases the assumption of treatment-block additivity \citep{Draper1998} is desirable. \cite{Gilmour2000PErsm} argue for the preservation of treatment-unit additivity when either fixed or random block effects are in the model in order to preserve consistency with the unblocked experiments; that would also conveniently imply treatment-block additivity. Adoption of these assumptions would then lead to the following representation of the relationship between responses and treatments in multi-stratum experiments.

@@

Based on the interrelation of the two models, and the desirability of being able to provide the possibility for testing for the lack-of-fit together with obtaining robust estimates of the variance components, the approach of using pure error is adopted, \cite{GilmourGoos2016Robust} explored several approaches to estimating the variance components. \\[5 pt]
[OE: the end of the part to be re-written/excluded]

The first approach is stratum-by-stratum data analysis, where each randomisation level is considered separately, and at each stratum $i$ units are treated as runs aggregated in $m_{i-1}$ blocks with fixed effects. The lower stratum variance is estimated from within blocks (using intra-block residual mean square), and the higher stratum (inter-block) variance from the difference between the intra- and inter-block residual mean squares scaled by the number of runs per block \citep{Hinkelmann2005Advanced}. This strategy requires only the treatment-unit additivity assumption and provides unbiased treatment effects' estimators. However, it would be desirable to make use of combining information not just from within-strata (intra-block) treatment replicates, but also from between-strata (inter-block) ones, especially in the common cases of relatively small experiments when the amount of runs available does not allow for obtaining higher stratum variance components from only considering blocks as whole units. Yates' procedure, first suggested by \cite{yates1939recovery} and described in detail by \cite{Hinkelmann2005Advanced}, provides more degrees of freedom for the variance estimate in the higher stratum -- and this is the approach we are adopting in this work.

In the case of two levels of randomisation, variance components are estimated in two variance decomposition steps: 
\begin{enumerate}
	\item Total SS = SS(Blocks) + SS(Treatments$\vert$Blocks) + SS(Residual)\\
	From fitting this full treatment-after-blocks model the residual mean square $S^2$ is taken as the estimate of the intra-block variance. SS(Residual) is then substituted in the following partition.
	\item Total SS = SS(Treatments) + SS(Blocks$\vert$Treatments) + SS(Residual)\\
	Therefore, the sums of squares for the blocks after the treatment effects have been accounted for is obtained and the corresponding mean square $$S^2_{b}=\mbox{SS(Blocks$\vert$Treatments)}/\nu_{b}$$ is then used to get an estimate of the inter-block variance:
	$\hat{\sigma}^2_{b}=\frac{\nu_{b}(S^2_{b}-S^2)}{d}$,
	where pure error degrees of freedom for inter-block variance is $d=n-\mbox{trace}[\bm{Z}'\bm{X_t}(\bm{X_t}'\bm{X_t})^{-1}\bm{X_t}'\bm{Z}]$, $\nu_{b}=\mbox{rank}([\bm{X_t} \bm{Z}])-\mbox{rank}(\bm{X_t})$, and $\bm{X_t}$ and $\bm{Z}$ are as in (\ref{eq::treatmentMS}).
\end{enumerate}

Allocation of the degrees of freedom in practice, following Yates' procedure, is provided in the Appendix.

[OE: to shorten]\\
\cite{GilmourGoos2016Robust} show that in addition to the treatment-unit additivity and randomisation the assumption of the experimental units being a random sample from an infinite population is necessary for the inter-block variance estimate to be unbiased. In this particular context, when the normality and independence of responses in (\ref{eq::MS_model}) is a standard assumption, although a potential presence of contamination effects in the fixed part of the model is also accounted for, Yates' approach seems to be the most appropriate technique. However, relying on the distribution assumption to the extent that makes REML the most sensible approach means that the fixed part of the fitted model is assumed to be absolutely true, that is the parameters of the population distribution are known, which does not comply with the model uncertainty framework.

@@

\subsection{MSE-based compound criteria for multistratum experiments}
\label{sec::ch7_search}
In the presence of potential model disturbance which is expressed as additional polynomial terms, the full model for a multistratum experiment is then:
\begin{equation}
\label{eq::MS_model_full}
\bm{Y}=\bm{X}_p\bm{\beta}_p+\bm{X}_q\bm{\beta}_q+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{X}_p$ is the primary terms matrix, and $bm{X}_q$ contains potential terms.

The design construction will follow the approach developed by \cite{Trinca2015improved}: an iterative, stratum-by-stratum algorithm, that treats the higher stratum units as fixed blocks; such an approach does not require any prior assumptions regarding the values of the variance components.

@@

\subsection*{Criteria for blocked experiments}
\label{subsec::compound_blocked}
\input{compound_blocked}

%Before moving to a more detailed presentation of the design construction algorithm, recall the MSE-based criteria (\ref{eq::MSE_D}), (\ref{eq::MSE_L}), (\ref{eq::MSE_D_B}) and (\ref{eq::MSE_L_B}) that are to be used:
%
%For an unblocked experiment:
%\begin{itemize}
%\item Determinant-based $MSE(D)$ criterion:
%\begin{align}
%\label{eq::MSE_D_ms}
%\mbox{minimise }&\left[\left|\bm{X}'_{1}\bm{X}_{1}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{X}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
%\end{align}
%\item Trace-based $MSE(L)$ criterion:
%\begin{align}
%\label{eq::MSE_L_ms}
%\mbox{minimise} &\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{X}_{1})^{-1}F_{1,d;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag\\& \left[\frac{1}{q}\mbox{trace}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times 
%\notag\\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{X}_{1})^{-1}+\tau^2\bm{A}\bm{A}'\}\right]^{\kappa_{MSE}}_{.}
%\end{align}
%\end{itemize} 
% 
%For a blocked experiment:
%\begin{itemize}
%\item Determinant-based $MSE(D)$ criterion:
%\begin{align}
%\label{eq::MSE_D_B_ms}
%\mbox{minimise }&\left[\left|(\bm{X}'_{1}\bm{Q}\bm{X}_{1})^{-1}\right|^{1/p}F_{p,d_B;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{QX}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{QX}_1\bm{\tilde{M}}^{-1}\bm{X}_1^{'}\bm{QX}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
%\end{align}
%\item Trace-based $MSE(L)$ criterion:
%\begin{align}
%\label{eq::MSE_L_B_ms}
%\mbox{minimise }&\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{Q}\bm{X}_{1})^{-1}F_{1,d_B;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag \\&\left[\frac{1}{q}\mbox{trace}\left(\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag \\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{QX}_{1})^{-1}+\tau^2[\bm{\tilde{A}}\bm{\tilde{A}}']_{22}\}\right]^{\kappa_{MSE}}_{.}
%\end{align}
%\end{itemize}

@@
